---
title: "Create custom line types"
author: "Dag Tanneberg"
date: "31 October 2015"
output: html_document
---

## Introduction

Line types are one very efficient way to map discrete
variables to the aesthetics of a plot. However, you
may quickly run out of alternatives if those variables have
more than a handful of possible values. Fortunately,
Hadley Wickham's ggplot2 supports custom line types, and the
default line types shown below merely represent quick
settings from a more complex set of alternatives.

```{r}
dta <- data.frame(
  line.type = c(
    "blank", "solid", "dashed", "dotted", "dotdash", 
    "longdash", "twodash"
  )
)

library('ggplot2')
ggplot(data = dta) +
  geom_segment(
    aes(
      x = 0, xend = 1, y = line.type, yend = line.type,
      linetype=line.type
    )
  ) +
  scale_y_discrete(name="linetype") +
  scale_linetype_identity() +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks = element_blank(),
    axis.title.x = element_blank(),
    panel.grid = element_blank()
  )
```

## Define a custom line type

To define your own line type you must provide a string
of length 2, 4, 6, or 8. The longer that string the more 
complex patterns become possible. Each element of the
vector tells ggplot2 about the width of fills and blanks in
a new line type. The pattern is FILL BLANK FILL BLANK
FILL BLANK FILL BLANK. Valid values range from 1 to 9 with
higher numbers indicating more width. The example below
plots 7 random line types of length 2.

```{r,fig.height=12}
dta <- expand.grid(1:9, 1:9)
dta[, 'line.type'] <- paste0(dta[, 1], dta[, 2])
head(dta)

set.seed(15889)
ggplot(data = dta[sample(1:nrow(dta), 7), ]) +
  geom_segment(
    aes(
      x = 0, xend = 1, y = line.type, yend = line.type,
      linetype=line.type
    )
  ) +
  scale_x_continuous(expand = c(0,0)) +
  scale_linetype_identity() +
  labs(y = 'Line type') +
  theme_minimal() +
  theme(
    axis.text.x = element_blank(),
    axis.ticks = element_blank(),
    axis.title.x = element_blank(),
    panel.grid = element_blank()
  )
```

## Note

Another introduction to custom line types is available from 
[Sape](http://sape.inf.usi.ch/quick-reference/ggplot2/linetype).
This resource mentions that ggplot2's line.type parameter
is equivalent to the base plot parameter lty. In other words,
you can do the same thing in base R.
