---
title: "Changing the graphical presentation of multiplicative
interaction effects: Improving the communication of
empirical social research"
author: "Dag Tanneberg"
data: "November 20, 2015"
output: 
  pdf_document:
    fig_width: 7
    fig_height: 5
---

# General considerations

Consider the standard linear model with a single
multiplicative interaction term. Let $x$ denote the
predictor of substantive interest and let $z$ be the
moderator. The fully specified regression fully 
$$
  y = f(x,z, \epsilon)
  y = \beta_0 + \beta_1 x + \beta_2 z + \beta_3 xz + \epsilon
$$

| $\beta_1$ | $\beta_2$ | $\beta_3$ | Joint Effect |
| --- | --- | --- | --- |
| > 0 | > 0 | > 0 | Mutual augmentation
| < 0 | > 0 | > 0 | $\beta_1$ attenuated, $\beta_2$ augmented
| > 0 | < 0 | > 0 | $\beta_1$ augmented, $\beta_2$ attenuated
| > 0 | > 0 | < 0 | Mutual attenuation
| < 0 | > 0 | < 0 | $\beta_1$ augmented, $\beta_2$ attenuated
| > 0 | < 0 | < 0 | $\beta_1$ attenuated, $\beta_2$ augmented
| < 0 | < 0 | < 0 | Mutual augmentation
| < 0 | < 0 | > 0 | Mutual attenuation

Assuming $x, z \ge 0$.

# Toy data setup

Original: Kam & Franzese, Multiplicative Interaction Terms
Download:
[Government duration data](http://www.press.umich.edu/resources/govduration.dta)

```{r, echo=FALSE}
# load data ------------------------------------------------
rm(list = ls())
library('ggplot2')
dta <- data.frame(
  id = c(
    2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16,
    17, 18, 19, 20, 21, 22, 23
  ),
  govdur = c(
    17.1, 32.2, 11.9, 12.3, 32.8, 30.0, 30.8, 18.0,
    20.2, 13.6, 11.0, 26.6, 31.8, 45.1, 30.3, 25.2, 11.5,
    29.2, 24.7, 43.6, 22.4, 30.4
  ),
  PS = c(
    52.0, 58.9, 59.0, 52.1, 54.3, 55.0, 76.4, 61.7,
    41.1, 57.4, 63.3, 55.4, 50.9, 69.3, 62.0, 47.6, 60.0,
    50.5, 47.2, 80.4, 59.1, 57.2
  ),
  NP = c(
    1.0, 2.2, 4.3, 2.7, 1.0, 1.0, 1.8, 3.0, 1.8, 2.9,
    1.1, 2.2, 1.6, 2.0, 3.3, 1.8, 1.6, 1.0, 1.4, 3.8, 1.7,
    1.0
  ),
  PD = c(
    0, 1, 0, 0, 1, 0, 1, 1, 1, 1, 0, 1, 1, 1, 1, 1, 0,
    0, 1, 1, 1, 1
  )
)
head(dta)
# Dependent variable: govdur = government duration
# Independent variables:
#  PS = Political strength
#  NP = Number of Parties
#  PD = ???
```
\newpage

# Fit model & generate predicted values

```{r}
# fit model & generate predictions ------------------------
fit <- lm(govdur ~ NP*PS + PD, data = dta)
summary(fit)

quantiles <- c(.05, .25, .5, .75, .95)
pred.dta <- expand.grid(
  # expand.grid covers all quantile combinations
  NP = quantile(dta[, 'NP'], quantiles),
  PS = dta[, 'PS'],
  PD = mean(dta[, 'PD'])
)
pred.y <- predict(
  fit, newdata = pred.dta, type = 'response',
  se.fit = TRUE, interval = 'confidence'
)
pdta <- data.frame(
  pred.dta, pred.y[['fit']], se.fit = pred.y[['se.fit']]
)
rm(quantiles, pred.y)
pdta <- within(pdta,
  NP.cat <- factor(NP, levels = sort(unique(NP)),
    labels = c(
      'Q[5] == 1', 'Q[25] == 1.175', 'Q[50] == 1.8',
      'Q[75] == 2.575', 'Q[95] == 3.775'
    )
  )
)
```
\newpage

\section{Presentation using Predicted Values}

An equally powerful presentation of interaction effects
focuses on the relationship between the response and
the focal predictor Such effect displays plot the predicted
values of the response against the focal predictor, while
fixing the moderator at select values. In effect, the
marginal effect on the focal predictor is replaced with the
expected level of the dependent variable itself, and the
conditional relationship can be interpreted in the metric of
the depedent variable itself. Using these graphs makes it
fairly easy to translate formal statistical models into the
substantive quantities that led to the formulation of the
conditional hypothesis in the first place.

This variant of effect displays becomes a particularly
effective means for communicating statistical results if at
least the the focal predictor is a continuous variable. In
the past many scholars resorted to either discretization or
focused on 


```{r, fig.align='center'}
ggplot(pdta, aes(x = PS, y = fit)) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = .6) +
  geom_line() +
  geom_rug(sides = 'b') +
  labs(
    x = 'Political Strength', y = 'Fitted Political Stability'
  ) +
  facet_grid(~NP.cat, labeller = label_parsed) +
  theme_minimal(base_size = 12, base_family = 'serif') +
  theme(panel.border = element_rect(size = .3, colour = 'grey65', fill = 'transparent'))
ggplot(pdta, aes(x = NP, y = fit)) +
  geom_pointrange(aes(ymin = lwr, ymax = upr)) +
  facet_grid(~PS)
```

# Version 2: Simple slopes approach

The conditional marginal effect of the focal predictor is
plotted against the moderator. This plots emphasizes
areas of statistical significance.

```{r}
beta <- coef(fit)
omega <- beta['PS'] + beta['NP:PS'] * dta[, 'NP']
se.fit <- sqrt(
  diag(vcov(fit))['PS'] + 
  2 * dta[, 'NP'] * vcov(fit)['PS', 'NP:PS'] + 
  (dta[, 'NP']^2) * diag(vcov(fit))['NP:PS']
)

upr <- omega + 1.96 * se.fit
lwr <- omega - 1.96 * se.fit
pdta.bauer <- data.frame(
  omega = omega,
  upr = upr,
  lwr = lwr,
  se.fit = se.fit,
  NP = dta[, 'NP']
)
pdta.bauer <- within(pdta.bauer, {
  signif <- ifelse(lwr < 0 & upr > 0, 1, 0)
  signif <- factor(signif, 0:1, c('Yes', 'No'))
  }
)
rm(beta, omega, se.fit, upr, lwr)
ggplot(
  data = pdta.bauer, 
  aes(x = NP, y = omega, ymin = lwr, ymax = upr, colour = signif)
) +
  geom_pointrange()
```

# Version 3: Chloropleth map

This map plots both independent variables against each other.
The expected value of the response is used to as fill color.
This plots ignores uncertainty.

```{r}
# simple map ----------------------------------------------
ggplot(data = pdta,
  aes(
    x = factor(format(NP, digits = 2)),
    y = factor(format(PS, digits = 3)),
    fill = fit #, alpha = 1/se.fit
  )
) +
  geom_tile(colour = 'white', size = .3) +
  scale_fill_gradient(low = '#2166ac', high = '#b2182b') +
  labs(
    x = 'Number of Cabinet Parties',
    y = 'Political Strength',
    fill = 'Fitted Government Duration'
  ) +
  theme_minimal() +
  theme(
    text = element_text(family = 'serif'),
    legend.key.height = grid::unit(.5, 'lines'),
    legend.position = 'top',
    legend.direction = 'horizontal',
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    plot.background = element_blank()
  )
```
